---
title: "PlaNET_epiphenotype_estimation"
author: "Melanie Smith"
date: "2025-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script I will take the raw DNAme data and estimate ancestry, cell type composition and gestational age using the PlaNET DNAme package.  

# Load Libraries

```{r include=FALSE}

# Install the IlluminaHumanMethylationEPICmanifest package and the annotation package
# BiocManager::install(c(
#   "IlluminaHumanMethylationEPICmanifest",
#   "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"
# ))

# Data manipulation
library(dplyr)
library(tidyr)
library(readr)
library(stringr)

# DNAme
library(minfi)
library(wateRmelon)

# Prediction
library(planet)
library(EpiDISH)

# Plotting
library(ggplot2)
library(RColorBrewer)

# load example data
# data("plBetas")
data("plCellCpGsThird")
```

# Set Directories and files

```{r include=FALSE} 
# Set project directory
projectDir <- "D:/VM_Projects/20250820_DNAme_PE" # needs the "windows style" directory path (D:)

metaDataDir <- file.path(projectDir, "metaData")

baseDir <- file.path(projectDir, "rawData")

outdir <- file.path(projectDir, "outDir")
if (!dir.exists("outdir")) dir.create("outdir")

# Set input files

import_samplesheet_file <- file.path(metaDataDir, "minfi_samplesheet.csv")
import_metadata_file <- file.path(metaDataDir, "patholgy.csv")

# Set output files
out_cell_estimate_rpc_file <- file.path(outdir, "cell_estimate_rpc.csv")

```

# Import metadata file

```{r}
# Read in the minfi style samplesheet
targets <- read_csv(import_samplesheet_file)

# Read in the metadata samplesheet
pathology <- read_csv(import_metadata_file)
```

# Read in the idat files

```{r}
# create the RGChannelSet object
RGset <- read.metharray.exp(
  base = baseDir,
  targets = targets
  )

# Get an overview of the data
# RGset
pData(RGset)
getManifest(RGset)

# Check the array type of your RGset
annotation(RGset)

```


## Remove samples not required for this analysis
None of the samples have failed the applied QC tests but I do have reservations about chip 203013220079. Fortunately these are all PAC samples so won't affect the PE analysis.  
Here I will retain only samples to be used in the PE analysis.  

```{r}

# Remove SCP3492 - this sample is from an EOPE pregnancy
keep <- !colnames(RGset) == "203013220099_R05C01"
# subset RGset
RGset <- RGset[,keep]
# Check the sample has been removed by looking at the number of colnames
RGset
# subset target as well
targets <- targets[keep,]


# Create vector of sample names where study equals "PE_placenta"
pe_samples <- targets$Basename[targets$study == "PE_placenta"]
# subset RGset
rgSet_pe <- RGset[,pe_samples]
# Check we have the expected number of samples
dim(rgSet_pe) # 105115 60

# create the phenoData for use downstream
phenoData_pe <- data.frame(colData(rgSet_pe)) %>%
  dplyr::mutate(samplename = gsub("_T$", "", Sample_Name) %>%
                           gsub("_", "", .)) %>%
  dplyr::left_join(., pathology, by = "samplename") %>%
  dplyr::select(., samplename, Sentrix_ID, Basename, cohort, Outcome, Sex, maternalAge, GestationalAge)


# subset target as well
targets_pe <- subset(targets, Basename %in% pe_samples)
dim(targets_pe) # 6 6
```

# Normalisation

```{r}
# Instead of preprocessQuantile, we will use preprocessNoob() followed by BMIQ.
# Note: At this stage we are keeping all probes (no filtering - but `preprocessNoob()` will perform some filtering automatically).

# Step 1: Background correction + dye bias normalization
mSetNoob_pe <- preprocessNoob(rgSet_pe)

# Step 2: Extract Beta values
betas_pe <- getBeta(mSetNoob_pe)

# Step 3: Get probe annotation aligned to Beta matrix
anno_pe <- getAnnotation(mSetNoob_pe)

# Step 4: Build design vector (1 = Type I, 2 = Type II) in correct order
design_vector_pe <- ifelse(anno_pe$Type == "I", 1, 2)

# Step 5: Apply BMIQ to each sample
bmiq_betas_pe <- apply(betas_pe, 2, function(sample_betas) {
  BMIQ(beta.v = sample_betas,
       design.v = design_vector_pe,
       plots = FALSE)$nbeta
})

# # Step 6: Replace Beta values in object
# assay(mSetNoob_pe, "Beta") <- bmiq_betas_pe
# mSetNoobBMIQ_pe <- mSetNoob_pe

# library(parallel)
# library(doParallel)
# library(foreach)
# 
# # Detect optimal cores (leave 1-2 free for system)
# n_cores <- max(1, detectCores() - 2)
# cat("Using", n_cores, "cores for BMIQ normalization\n")
# 
# # Set up parallel processing
# cl <- makeCluster(n_cores)
# registerDoParallel(cl)
# clusterExport(cl, c("BMIQ", "design_vector_pe"), envir = environment())
# 
# # Parallel BMIQ
# bmiq_betas_pe <- foreach(i = 1:ncol(betas_pe),
#                         .packages = c("wateRmelon"),
#                         .combine = cbind,
#                         .errorhandling = "pass") %dopar% {
# 
#   BMIQ(beta.v = betas_pe[, i],
#        design.v = design_vector_pe,
#        plots = FALSE)$nbeta
# }
# 
# # Cleanup
# stopCluster(cl)
# registerDoSEQ()
# 
# # Restore column/row names
# colnames(bmiq_betas_pe) <- colnames(betas_pe)
# rownames(bmiq_betas_pe) <- rownames(betas_pe)
```

# Reference-based: planet
planet uses a placenta-specific reference panel (trophoblast, Hofbauer, stromal, endothelial cells) to deconvolve cell proportions from Illumina 450k/EPIC arrays. 

## Minfi

```{r}
placenta_estimates_pe <- minfi:::projectCellType(
  bmiq_betas_pe[rownames(plCellCpGsThird), ],
  plCellCpGsThird,
  lessThanOne = FALSE
  )

head(placenta_estimates_pe)
```

## EpiDISH

```{r}
# robust partial correlations -> use this one (see lab book for notes)
epidish_RPC <- epidish(
  beta.m = bmiq_betas_pe[rownames(plCellCpGsThird), ],
  ref.m = plCellCpGsThird,
  method = "RPC"
)
estimates_RPC <- epidish_RPC$estF %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Basename") %>%
  dplyr::left_join(., phenoData_pe, by = "Basename")

# write_csv(estimates_RPC,
#           file = file.path(out_cell_estimate_rpc_file))

# CIBERSORT
epidish_CBS <- epidish(
  beta.m = bmiq_betas_pe[rownames(plCellCpGsThird), ],
  ref.m = plCellCpGsThird,
  method = "CBS"
)

# constrained projection (houseman 2012)
epidish_CP <- epidish(
  beta.m = bmiq_betas_pe[rownames(plCellCpGsThird), ],
  ref.m = plCellCpGsThird,
  method = "CP"
)

```

# Compare

```{r}
data("plColors")

# bind estimate data frames and reshape for plotting
bind_rows(
  placenta_estimates_pe %>% as.data.frame() %>% mutate(algorithm = "CP (placenta_estimates_pe)"),
  epidish_RPC$estF %>% as.data.frame() %>% mutate(algorithm = "RPC"),
  epidish_CBS$estF %>% as.data.frame() %>% mutate(algorithm = "CBS"),
  epidish_CP$estF %>% as.data.frame() %>% mutate(algorithm = "CP (EpiDISH)")
) %>%
  mutate(sample = rep(rownames(placenta_estimates_pe), 4)) %>%
  as_tibble() %>%
  pivot_longer(
    cols = -c(algorithm, sample),
    names_to = "component",
    values_to = "estimate"
  ) %>%
  mutate(component = factor(component,
                            levels = c("nRBC", "Endothelial", "Hofbauer",
                                       "Stromal", "Trophoblasts", "Syncytiotrophoblast")
  )) %>%
  dplyr::left_join(
    phenoData_pe %>% dplyr::select(Basename, samplename, Outcome, GestationalAge),
    by = c("sample" = "Basename")
  ) %>%
  ## 🔑 reorder factor levels for x-axis
  mutate(samplename = forcats::fct_reorder(samplename, GestationalAge, .desc = FALSE)) %>%
  
  # plot
  ggplot(aes(x = samplename, y = estimate, fill = component)) +
  geom_bar(stat = "identity") +
  facet_wrap(~algorithm, ncol = 1) +
  scale_fill_manual(values = plColors) +
  scale_y_continuous(
    limits = c(-0.1, 1.1), breaks = c(0, 0.5, 1),
    labels = scales::percent
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "", fill = "")

```

# Gestational Age Prediction
There are 3 gestational age clocks for placental DNA methylation data from ([Lee 2019](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6628997/)):

Robust Placental Clock (RPC)
Control Placental Clock (CPC)
Refined Robust Placental Clock (RRPC) -> use this one (see lab book for notes)  

```{r}

phenoData_pe_ga <- phenoData_pe %>%
  mutate(
    ga_RPC = predictAge(bmiq_betas_pe, type = "RPC"),
    ga_CPC = predictAge(bmiq_betas_pe, type = "CPC"),
    ga_RRPC = predictAge(bmiq_betas_pe, type = "RRPC")
  )

# 558 of 558 predictors present.
# 546 of 546 predictors present.
# 395 of 395 predictors present.

# plot the predicted gestational age as a function of reported gestational age, stratified by outcome
phenoData_pe_ga %>%

  # reshape, to plot
  pivot_longer(
    cols = contains("ga"),
    names_to = "clock_type",
    names_prefix = "ga_",
    values_to = "ga"
  ) %>%
  
  # plot code
  ggplot(aes(x = GestationalAge, y = ga, col = Outcome)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~clock_type) +
  theme(legend.position = "top") +
  labs(x = "Reported GA (weeks)", y = "Inferred GA (weeks)", col = "")

# plot the predicted gestational age as a function of reported gestational age stratified by sex and outcome
phenoData_pe_ga %>%
  dplyr::mutate(.,
                sex_outcome = paste0(Sex, "_", Outcome)) %>%

  # reshape, to plot
  pivot_longer(
    cols = contains("ga"),
    names_to = "clock_type",
    names_prefix = "ga_",
    values_to = "ga"
  ) %>%

ggplot(aes(x = GestationalAge, y = ga, col = sex_outcome)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "black") +
  facet_wrap(~clock_type) +
  coord_equal() +
  labs(x = "Reported GA (weeks)", y = "Inferred GA (weeks)", col = "") + 
  xlim(35, 43) + ylim(35, 43) +
  theme_bw() +
    theme(legend.position = "top")

```

# Predict ethnicity

```{r}

data(ethnicityCpGs)
all(ethnicityCpGs %in% rownames(betas_pe))
all(ethnicityCpGs %in% rownames(bmiq_betas_pe))

results_ethnicity <- predictEthnicity(bmiq_betas_pe, force = TRUE)

# Compare the predicted ethnicity with self report
compare_ethnicity <- dplyr::left_join(
  results_ethnicity, phenoData_pe, 
  by = c("Sample_ID" = "Basename")
  ) %>%
  dplyr::left_join(., pathology,
                   by = "samplename") %>%
  dplyr::select(.,  samplename, self_report = Ethnicity, Predicted_ethnicity, Prob_African, Prob_Asian, Prob_Caucasian)



```

## Understanding which probes are being removed

```{r}
# Check dimensions at each step
cat("Original RGset probes:", nrow(rgSet_pe), "\n")
cat("After preprocessNoob:", nrow(mSetNoob_pe), "\n")
cat("After getBeta:", nrow(betas_pe), "\n")
cat("After BMIQ:", nrow(bmiq_betas_pe), "\n")

# Check for NA introduction
na_before_bmiq <- sum(is.na(betas_pe))
na_after_bmiq <- sum(is.na(bmiq_betas_pe))
cat("NAs before BMIQ:", na_before_bmiq, "\n")
cat("NAs after BMIQ:", na_after_bmiq, "\n")

# Check specific ethnicity probes
data(ethnicityCpGs)
in_original <- sum(ethnicityCpGs %in% rownames(betas_pe))
in_bmiq <- sum(ethnicityCpGs %in% rownames(bmiq_betas_pe))
cat("Ethnicity probes before BMIQ:", in_original, "/", length(ethnicityCpGs), "\n")
cat("Ethnicity probes after BMIQ:", in_bmiq, "/", length(ethnicityCpGs), "\n")

# check which 15 probes are missing

missing_probes <- setdiff(ethnicityCpGs, rownames(betas_pe))
length(missing_probes)
missing_probes

# are these probes missing from the 850k Epic array platform?
missing_probes %in% rownames(anno_pe)


```

I can see that the missing probes are rs... IDs (SNP probes, not CpG probes).  
On the Illumina arrays:  

- Most methylation probes have names like cg00000029, cg18478105, etc.  
- Some probes are designed to genotype SNPs instead of measuring methylation — those are labeled rs....  

Why they’re missing in your data
On EPIC (850k) arrays, many of the older SNP control probes that existed on 450K arrays were dropped.
That’s why your anno_pe (from EPIC) has only cg... IDs, and none of those rs... SNP IDs.
The ethnicityCpGs reference set was built originally for 450K and still contains SNP-based predictors.
So this is not a naming mismatch — the probes simply aren’t present on the EPIC array you’re using.
How this affects predictEthnicity()
The function expects 1860 predictors, but with EPIC you only have 1845 (the cg... probes).
Fortunately, the SNP probes are not strictly required — the model will usually still work if you pass a matrix that contains them as NA.

# CYT:SYN ratio Wilcoxon Rank Sum Test

```{r}
# Calculate the cyt:syn ratio
estimates_RPC$cyt_syn_ratio <- estimates_RPC$Trophoblasts / estimates_RPC$Syncytiotrophoblast

# Check the distribution
summary(estimates_RPC$cyt_syn_ratio)
hist(estimates_RPC$cyt_syn_ratio, main = "Cytotrophoblast:Syncytiotrophoblast Ratio")

# Wilcoxon Rank Sum Tests as specified

# 1. Test by fetal sex
wilcox_sex <- wilcox.test(cyt_syn_ratio ~ Sex, 
                         data = estimates_RPC,
                         exact = FALSE)
print("Cyt:Syn ratio by fetal sex:")
print(wilcox_sex)

# plot cyt:syn by fetal sex
estimates_RPC %>%
  
  ggplot(., aes(x = Sex,
                y = cyt_syn_ratio,
                colour = Sex,
                fill = Sex)) +
  
  geom_boxplot(alpha = 0.7,
               size = 0.8) +  # Darker outline, transparent fill
  
  geom_jitter(width = 0.05,
              alpha = 0.6,
              size = 2) +  # Add jittered points
  
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male")) +  # Rename x-axis labels
  
  labs(x = "Fetal Sex", y = "Cyt:Syn") +  # Add axis titles
  
  theme_bw() +
    theme(
    legend.position = "none",  # Remove legend since x-axis is self-explanatory
    
    axis.text = element_text(size = 14,
                             color = "black"),    # Larger, black axis labels
    
    axis.title = element_text(size = 16,
                              color = "black")    # Larger, black axis titles
  )


# 2. Test total trophoblast by fetal sex

# Combine the cyt & syn
estimates_RPC$total_trophoblast <- estimates_RPC$Trophoblasts + estimates_RPC$Syncytiotrophoblast

wilcox_total_sex <- wilcox.test(total_trophoblast ~ Sex, 
                         data = estimates_RPC,
                         exact = FALSE)
print("Total trophoblast by fetal sex:")
print(wilcox_total_sex)

# plot total trophoblasts by fetal sex
estimates_RPC %>%
  
  ggplot(., aes(x = Sex,
                y = total_trophoblast,
                colour = Sex,
                fill = Sex)) +
  
  geom_boxplot(alpha = 0.7,
               size = 0.8) +  # Darker outline, transparent fill
  
  geom_jitter(width = 0.05,
              alpha = 0.6,
              size = 2) +  # Add jittered points
  
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male")) +  # Rename x-axis labels
  
  labs(x = "Fetal Sex", y = "Total Trophoblasts") +  # Add axis titles
  
  theme_bw() +
    theme(
    legend.position = "none",  # Remove legend since x-axis is self-explanatory
    
    axis.text = element_text(size = 14,
                             color = "black"),    # Larger, black axis labels
    
    axis.title = element_text(size = 16,
                              color = "black")    # Larger, black axis titles
  )

# 3. Test combined Endothelial and Stromal by fetal sex

# Combine the endothelial and stromal cells
estimates_RPC$combined_endo_stromal <- estimates_RPC$Endothelial + estimates_RPC$Stromal

wilcox_endo_stromo_sex <- wilcox.test(combined_endo_stromal ~ Sex, 
                         data = estimates_RPC,
                         exact = FALSE)
print("Combined Endothelial and Stromal by fetal sex:")
print(wilcox_total_sex)

# plot  combined Endothelial and Stromal by fetal sex
estimates_RPC %>%
  
  ggplot(., aes(x = Sex,
                y = combined_endo_stromal,
                colour = Sex,
                fill = Sex)) +
  
  geom_boxplot(alpha = 0.7,
               size = 0.8) +  # Darker outline, transparent fill
  
  geom_jitter(width = 0.05,
              alpha = 0.6,
              size = 2) +  # Add jittered points
  
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male")) +  # Rename x-axis labels
  
  labs(x = "Fetal Sex", y = "Combined Endothelial and Stromal") +  # Add axis titles
  
  theme_bw() +
    theme(
    legend.position = "none",  # Remove legend since x-axis is self-explanatory
    
    axis.text = element_text(size = 14,
                             color = "black"),    # Larger, black axis labels
    
    axis.title = element_text(size = 16,
                              color = "black")    # Larger, black axis titles
  )

# 4. Test combined nucleated RBCs and Hofbauer cells by fetal sex

# Combine the nucleated RBCs and Hofbauer cells
estimates_RPC$combined_nRBC_hof <- estimates_RPC$nRBC + estimates_RPC$Hofbauer

wilcox_nRBC_hofsex <- wilcox.test(combined_nRBC_hof ~ Sex, 
                         data = estimates_RPC,
                         exact = FALSE)
print("Combined nucleated RBCs and Hofbauer cells by fetal sex:")
print(wilcox_total_sex)

# plot  combined nucleated RBCs and Hofbauer cells by fetal sex
estimates_RPC %>%
  
  ggplot(., aes(x = Sex,
                y = combined_endo_stromal,
                colour = Sex,
                fill = Sex)) +
  
  geom_boxplot(alpha = 0.7,
               size = 0.8) +  # Darker outline, transparent fill
  
  geom_jitter(width = 0.05,
              alpha = 0.6,
              size = 2) +  # Add jittered points
  
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male")) +  # Rename x-axis labels
  
  labs(x = "Fetal Sex", y = "Combined nRBCs and Hofbauer cells") +  # Add axis titles
  
  theme_bw() +
    theme(
    legend.position = "none",  # Remove legend since x-axis is self-explanatory
    
    axis.text = element_text(size = 14,
                             color = "black"),    # Larger, black axis labels
    
    axis.title = element_text(size = 16,
                              color = "black")    # Larger, black axis titles
  )

# 5. Test by outcome (PE vs Control)
wilcox_outcome <- wilcox.test(cyt_syn_ratio ~ Outcome, 
                             data = estimates_RPC,
                             exact = FALSE)
print("Cyt:Syn ratio by outcome:")
print(wilcox_outcome)

# 6. For gestational age - since it's continuous, you might want to:
# Test correlation for all samples
cor_ga <- cor.test(estimates_RPC$cyt_syn_ratio, 
                  estimates_RPC$GestationalAge, 
                  method = "spearman")
print("Cyt:Syn ratio correlation with gestational age:")
print(cor_ga)

# Test correlation sex stratified
# Male correlation
cor_ga_male <- cor.test(estimates_RPC$cyt_syn_ratio[estimates_RPC$Sex == "M"], 
                       estimates_RPC$GestationalAge[estimates_RPC$Sex == "M"], 
                       method = "spearman")

# Female correlation
cor_ga_female <- cor.test(estimates_RPC$cyt_syn_ratio[estimates_RPC$Sex == "F"], 
                         estimates_RPC$GestationalAge[estimates_RPC$Sex == "F"], 
                         method = "spearman")

# Print results
print("Male correlation:")
print(cor_ga_male)

print("Female correlation:")
print(cor_ga_female)

# Summary statistics by groups
aggregate(cyt_syn_ratio ~ Sex, data = estimates_RPC, 
          FUN = function(x) c(median = median(x), IQR = IQR(x)))
aggregate(cyt_syn_ratio ~ Outcome, data = estimates_RPC, 
          FUN = function(x) c(median = median(x), IQR = IQR(x)))

# Correlation plot with trend lines by sex
estimates_RPC %>%
  ggplot(aes(
    x = GestationalAge,
    y = cyt_syn_ratio,
    colour = Sex
    )) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add trend lines with confidence intervals
  # Use non-parametric smoothing instead of linear
# geom_smooth(method = "loess", se = TRUE)
  scale_colour_discrete(name = "Fetal Sex", 
                       labels = c("F" = "Female", "M" = "Male")) +  # Rename legend labels
  labs(x = "Gestational Age at Delivery (weeks)", y = "Cyt:Syn Ratio") +  # Corrected axis titles
  theme_bw() +
  theme(
    legend.position = "bottom",  # Show legend at bottom
    axis.text = element_text(size = 14, color = "black"),    
    axis.title = element_text(size = 16, color = "black"),
    legend.text = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 14, color = "black")
  )
```

## Statistical difference - gestational age by sex

```{r}
# Split by sex
males <- estimates_RPC$GestationalAge[estimates_RPC$Sex == "M"]
females <- estimates_RPC$GestationalAge[estimates_RPC$Sex == "F"]

print("=== SAMPLE SIZES ===")
cat("Males:", length(males), "\n")
cat("Females:", length(females), "\n")

print("\n=== DESCRIPTIVE STATISTICS ===")
# Summary statistics
stats_summary <- estimates_RPC %>%
  group_by(Sex) %>%
  summarise(
    n = n(),
    mean = round(mean(GestationalAge), 2),
    median = round(median(GestationalAge), 2),
    sd = round(sd(GestationalAge), 2),
    min = round(min(GestationalAge), 2),
    max = round(max(GestationalAge), 2),
    .groups = 'drop'
  )
print(stats_summary)

# 1. Test for normality
print("Shapiro-Wilk normality tests:")
shapiro_male <- shapiro.test(males)
shapiro_female <- shapiro.test(females)

cat("Males: W =", round(shapiro_male$statistic, 4), 
    ", p-value =", ifelse(shapiro_male$p.value < 0.001, "<0.001", round(shapiro_male$p.value, 4)), "\n")
cat("Females: W =", round(shapiro_female$statistic, 4), 
    ", p-value =", ifelse(shapiro_female$p.value < 0.001, "<0.001", round(shapiro_female$p.value, 4)), "\n")

# 2. Test for equal variances
levene_test <- var.test(males, females)
cat("\nF-test for equal variances: F =", round(levene_test$statistic, 4),
    ", p-value =", ifelse(levene_test$p.value < 0.001, "<0.001", round(levene_test$p.value, 4)), "\n")

# Determine normality
normal_males <- shapiro_male$p.value > 0.05
normal_females <- shapiro_female$p.value > 0.05
both_normal <- normal_males & normal_females
equal_vars <- levene_test$p.value > 0.05

cat("\nAssumption check:")
cat("\n- Males normally distributed:", normal_males)
cat("\n- Females normally distributed:", normal_females)
cat("\n- Equal variances:", equal_vars)

print("\n=== STATISTICAL TESTS ===")

# Parametric test (t-test)
if(both_normal) {
  if(equal_vars) {
    t_test <- t.test(males, females, var.equal = TRUE)
    cat("Using: Student's t-test (equal variances)\n")
  } else {
    t_test <- t.test(males, females, var.equal = FALSE)
    cat("Using: Welch's t-test (unequal variances)\n")
  }
  
  cat("t =", round(t_test$statistic, 4))
  cat(", df =", round(t_test$parameter, 1))
  cat(", p-value =", ifelse(t_test$p.value < 0.001, "<0.001", round(t_test$p.value, 4)))
  cat("\nMean difference =", round(t_test$estimate[1] - t_test$estimate[2], 3), "weeks")
  cat("\n95% CI:", round(t_test$conf.int[1], 3), "to", round(t_test$conf.int[2], 3), "\n")
  
} else {
  cat("Normality assumption violated - t-test not recommended\n")
}

# Non-parametric test (Mann-Whitney U / Wilcoxon rank-sum)
wilcox_test <- wilcox.test(males, females, conf.int = TRUE)
cat("\nMann-Whitney U test (non-parametric):")
cat("\nW =", wilcox_test$statistic)
cat(", p-value =", ifelse(wilcox_test$p.value < 0.001, "<0.001", round(wilcox_test$p.value, 4)))
cat("\nMedian difference =", round(wilcox_test$estimate, 3), "weeks")
cat("\n95% CI:", round(wilcox_test$conf.int[1], 3), "to", round(wilcox_test$conf.int[2], 3), "\n")

print("\n=== RECOMMENDATION ===")
if(both_normal) {
  cat("✓ Use parametric test (t-test) - normality assumptions met\n")
  if(equal_vars) {
    cat("✓ Use Student's t-test (equal variances)\n")
  } else {
    cat("✓ Use Welch's t-test (unequal variances)\n")
  }
} else {
  cat("✓ Use non-parametric test (Mann-Whitney U) - normality violated\n")
}

# Effect size (Cohen's d)
pooled_sd <- sqrt(((length(males)-1)*var(males) + (length(females)-1)*var(females)) / 
                  (length(males) + length(females) - 2))
cohens_d <- (mean(males) - mean(females)) / pooled_sd
cat("\nCohen's d (effect size) =", round(cohens_d, 3))

effect_interpretation <- case_when(
  abs(cohens_d) < 0.2 ~ "negligible",
  abs(cohens_d) < 0.5 ~ "small", 
  abs(cohens_d) < 0.8 ~ "medium",
  TRUE ~ "large"
)
cat(" (", effect_interpretation, " effect)\n")

print("\n=== VISUALIZATION ===")
# Create comparison plots
p1 <- ggplot(estimates_RPC, aes(x = Sex, y = GestationalAge, fill = Sex)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male")) +
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue")) +
  labs(x = "Fetal Sex", y = "Gestational Age (weeks)", 
       title = "Gestational Age by Fetal Sex") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))

p2 <- ggplot(estimates_RPC, aes(x = GestationalAge, fill = Sex)) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 20) +
  scale_fill_manual(values = c("F" = "pink", "M" = "lightblue"),
                    labels = c("F" = "Female", "M" = "Male")) +
  labs(x = "Gestational Age (weeks)", y = "Count", 
       title = "Distribution of Gestational Age by Sex") +
  theme_bw() +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 14))

print("Box plot:")
print(p1)
print("Histogram:")  
print(p2)
```


# Metadata table

```{r}
# Demographics Table Generator
# Split by Sex (Male vs Female)

library(dplyr)
library(knitr)
library(kableExtra)

# Function to calculate mean ± SD
mean_sd <- function(x, digits = 2) {
  if(all(is.na(x))) return("NA")
  mean_val <- round(mean(x, na.rm = TRUE), digits)
  sd_val <- round(sd(x, na.rm = TRUE), digits)
  return(paste0(mean_val, " ± ", sd_val))
}

# Function to calculate n (%)
n_percent <- function(x, category = NULL, digits = 1) {
  if(is.null(category)) {
    n <- sum(!is.na(x))
    total <- length(x)
  } else {
    n <- sum(x == category, na.rm = TRUE)
    total <- sum(!is.na(x))
  }
  if(total == 0) return("0 (0.0)")
  percent <- round((n/total) * 100, digits)
  return(paste0(n, " (", percent, ")"))
}

# Create demographics table
create_demographics_table <- function(data) {
  
  # Split data by sex
  males <- data[data$Sex == "M", ]
  females <- data[data$Sex == "F", ]
  total <- data
  
  # Initialize results dataframe
  demo_table <- data.frame(
    Characteristic = character(),
    Male = character(),
    Female = character(),
    Total = character(),
    stringsAsFactors = FALSE
  )
  
  # Sample size
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Sample size, n",
    Male = nrow(males),
    Female = nrow(females),
    Total = nrow(total)
  ))
  
  # Gestational Age
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Gestational age at delivery, weeks (mean ± SD)",
    Male = mean_sd(males$GestationalAge),
    Female = mean_sd(females$GestationalAge),
    Total = mean_sd(total$GestationalAge)
  ))
  
  # Delivery Mode
  delivery_modes <- unique(c(males$deliveryMode, females$deliveryMode))
  delivery_modes <- delivery_modes[!is.na(delivery_modes)]
  
  for(mode in delivery_modes) {
    demo_table <- rbind(demo_table, data.frame(
      Characteristic = paste0("Delivery mode - ", mode, ", n (%)"),
      Male = n_percent(males$deliveryMode, mode),
      Female = n_percent(females$deliveryMode, mode),
      Total = n_percent(total$deliveryMode, mode)
    ))
  }
  
  # Preeclampsia
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Preeclampsia, n (%)",
    Male = n_percent(males$Outcome, "PE"),
    Female = n_percent(females$Outcome, "PE"),
    Total = n_percent(total$Outcome, "PE")
  ))
  
  # Maternal Ethnicity
  ethnicities <- unique(c(males$Ethnicity, females$Ethnicity))
  ethnicities <- ethnicities[!is.na(ethnicities)]
  
  for(eth in ethnicities) {
    demo_table <- rbind(demo_table, data.frame(
      Characteristic = paste0("Maternal ethnicity - ", eth, ", n (%)"),
      Male = n_percent(males$Ethnicity, eth),
      Female = n_percent(females$Ethnicity, eth),
      Total = n_percent(total$Ethnicity, eth)
    ))
  }
  
  # Maternal Age
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Maternal age, years (mean ± SD)",
    Male = mean_sd(males$maternalAge),
    Female = mean_sd(females$maternalAge),
    Total = mean_sd(total$maternalAge)
  ))
  
  # Maternal BMI
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Maternal BMI, kg/m² (mean ± SD)",
    Male = mean_sd(males$BMI),
    Female = mean_sd(females$BMI),
    Total = mean_sd(total$BMI)
  ))
  
  # Birthweight
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Birthweight, grams (mean ± SD)",
    Male = mean_sd(males$Birthweight),
    Female = mean_sd(females$Birthweight),
    Total = mean_sd(total$Birthweight)
  ))
  
  # Smoking Status
  demo_table <- rbind(demo_table, data.frame(
    Characteristic = "Smoking during pregnancy, n (%)",
    Male = n_percent(males$SmokingStatus, "Y"),
    Female = n_percent(females$SmokingStatus, "Y"),
    Total = n_percent(total$SmokingStatus, "Y")
  ))
  
  return(demo_table)
}

# Generate the table
demographics_table <- create_demographics_table(dplyr::filter(pathology, samplename != "SCP3492"))

# Display as a formatted table
kable(demographics_table, 
      col.names = c("Characteristic", "Male", "Female", "Total"),
      align = c("l", "c", "c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "4in") %>%
  column_spec(2:4, width = "1.2in")

# For publication-ready output, you can also export to different formats:
write.csv(demographics_table, 
          file.path(outdir, "demographics_table.csv"),
          row.names = FALSE)

# Print basic table to console
print(demographics_table)
```

# Session Info

```{r}
sessionInfo()
```
